---
title: "Guide for Model Developers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guide for Model Developers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

This guide explains how to make your model available for independent external
validation using evaluatr. Independent evaluators will be able to run
`secure_model_validation()` in their own datasets, without you ever seeing
their patient data and without them ever seeing your model's coefficients.

You will need:

1. A **private GitHub repository** to host the model specification file
2. The **evaluatr package** installed locally to generate that file
3. A **developer account** — just your GitHub username

This takes approximately 30 minutes to set up for the first model, and much
less for subsequent ones.

---

## Step 1: Create a private GitHub repository

If you do not already have a GitHub account, create one at
[github.com](https://github.com). Then create a new repository:

- Set the visibility to **Private**
- Choose a clear, memorable name (e.g. `my-clinical-models`)
- Do not initialise with a README

Your repository will hold one or more model specifications. The required
folder structure is:

```
my-clinical-models/
├── model_v1/
│   └── model_v1_specification.json
└── model_v2/
    └── model_v2_specification.json
```

The folder name (e.g. `model_v1`) becomes the `model_id` that evaluators
will use. The specification file must be named `{model_id}_specification.json`.

---

## Step 2: Install evaluatr and generate your model specification

The `register_model()` function creates the protected model specification
file and registers the model with the evaluatr registry. Both steps happen
in a single call and must complete before any evaluator can validate the model.

```{r}
devtools::install_github("JoieEnsor/evaluatr")
library(evaluatr)
```

### Example A: Logistic regression (DVT diagnosis model)

The DVT example dataset included with the package spans seven studies. Studies
1-5 are used here for model development; studies 6-7 are reserved for
external validation.

```{r}
dvt_path <- system.file("extdata", "dvt_example.csv", package = "evaluatr")
dvt_data <- read.csv(dvt_path)

# Development set: studies 1-5
dvt_dev <- dvt_data[dvt_data$study %in% 1:5, ]

# Fit a logistic regression model
dvt_fit <- glm(dvt ~ male + ab_ddimer + calf_diff + vein_distension +
                     malignancy + immobile,
               data = dvt_dev, family = binomial)

# Generate the protected model specification file
register_model(
  model           = dvt_fit,
  model_id        = "dvt_v1",
  developer_id    = "your-github-username",    # your GitHub username
  model_name      = "DVT Diagnosis Model v1",
  description     = paste(
    "Logistic regression for DVT diagnosis. Developed in studies 1-5",
    "of the simulated DVT example dataset.",
    "Predictors: male, ab_ddimer, calf_diff, vein_distension, malignancy, immobile."
  ),
  developer_name  = "Your Name",
  developer_email = "you@institution.ac.uk",
  output_dir      = "dvt_v1"
)
```

This writes `dvt_v1/dvt_v1_specification.json`. The coefficient values are
protected; the file is safe to host in your GitHub repository.

### Example B: Cox proportional hazards model (breast cancer relapse)

```{r}
library(survival)

bc_path <- system.file("extdata", "bc_survival_example.csv",
                       package = "evaluatr")
bc_data <- read.csv(bc_path)
bc_dev  <- bc_data[bc_data$val == 0, ]

# Fit a Cox model
cox_fit <- coxph(Surv(rf, rfi) ~ age + nodes + pr + er + hormon,
                 data = bc_dev, x = TRUE)

# Extract baseline survival at 5 and 10 years (60 and 120 months)
bh     <- basehaz(cox_fit, centered = FALSE)
S0_60  <- exp(-bh$hazard[which.min(abs(bh$time - 60))])
S0_120 <- exp(-bh$hazard[which.min(abs(bh$time - 120))])

register_model(
  model            = cox_fit,
  model_id         = "bc_cox_v1",
  developer_id     = "your-github-username",
  model_name       = "Breast Cancer Relapse-Free Survival Model",
  description      = paste(
    "Cox PH model for relapse-free survival in breast cancer.",
    "Developed in the training set (val == 0) of the bc_survival_example dataset.",
    "Predictors: age, nodes, pr, er, hormon."
  ),
  model_parameters = list(
    timepoints        = c(60, 120),
    baseline_survival = c(S0_60, S0_120)
  ),
  developer_name  = "Your Name",
  developer_email = "you@institution.ac.uk",
  output_dir      = "bc_cox_v1"
)
```

For Cox models, `model_parameters` must include `timepoints` and
`baseline_survival` so that evaluatr can produce predicted survival
probabilities at specific time-points.

### Example C: Model with embedded pre-processing

If your model uses derived or transformed predictors, you can embed the
transformation so that evaluators provide raw data with no additional steps
required on their side. Supply an R code string as the `preprocessing`
argument:

```{r}
# A Cox model using log-transformed pr and er
cox_log_fit <- coxph(Surv(rf, rfi) ~ age + nodes + log(pr + 1) + log(er + 1),
                     data = bc_dev, x = TRUE)

bh2     <- basehaz(cox_log_fit, centered = FALSE)
S0_60b  <- exp(-bh2$hazard[which.min(abs(bh2$time - 60))])
S0_120b <- exp(-bh2$hazard[which.min(abs(bh2$time - 120))])

register_model(
  model            = cox_log_fit,
  model_id         = "bc_cox_log_v1",
  developer_id     = "your-github-username",
  model_name       = "BC Survival Model (log biomarkers)",
  preprocessing    = "
    validation_data$`log(pr + 1)` <- log(validation_data$pr + 1)
    validation_data$`log(er + 1)` <- log(validation_data$er + 1)
  ",
  model_parameters = list(
    timepoints        = c(60, 120),
    baseline_survival = c(S0_60b, S0_120b)
  ),
  developer_name  = "Your Name",
  developer_email = "you@institution.ac.uk",
  output_dir      = "bc_cox_log_v1"
)
```

The preprocessing code is run on `validation_data` (the evaluator's dataset)
before prediction. The code must create any derived columns in place. It has
access only to `validation_data`.

If you do not embed preprocessing, the variable names in `variables` (or
the coefficient names extracted from your model) must match the column names
the evaluator will have in their dataset. Document any required
transformations clearly in the model description.

### Example D: Manual coefficient specification

If your coefficients come from a publication or software other than R, you
can supply them directly:

```{r}
register_model(
  coefficients = c("(Intercept)" = -3.2,
                   ab_ddimer     =  1.8,
                   calf_diff     =  1.1,
                   vein_distension = 0.9),
  model_type   = "logistic",
  model_id     = "dvt_published_v1",
  developer_id = "your-github-username",
  model_name   = "Published DVT Score (manual)",
  outcome_type = "binary",
  variables    = c("ab_ddimer", "calf_diff", "vein_distension"),
  description  = "Coefficients from Smith et al. (2024), Table 2.",
  developer_name  = "Your Name",
  developer_email = "you@institution.ac.uk",
  output_dir   = "dvt_published_v1"
)
```

### Keeping a model private while a paper is under review

Set `public_listing = FALSE` to register the model without it appearing in
the public directory. You can still issue tokens to peer reviewers or journal
editors. Update to `public_listing = TRUE` by re-running
`register_model()` with a new `model_id` when ready to publish.

---

## Step 3: Verify your model specification

Before sharing with any evaluator, confirm that evaluatr's predictions match
your own. Because `secure_model_validation()` returns shuffled predictions,
sort both sets before comparing:

```{r}
# Validate in your own development data using your own full-access PAT
result_check <- secure_model_validation(
  repo_owner      = "your-github-username",
  repo_name       = "my-clinical-models",
  model_id        = "dvt_v1",
  github_token    = Sys.getenv("GITHUB_PAT"),   # your own full-access PAT
  validation_data = dvt_dev,
  outcome         = "dvt"
)

# Your own predictions (unsorted)
manual_probs <- predict(dvt_fit, type = "response")

# Sort both and compare
max(abs(sort(manual_probs) - sort(result_check$shuffled_predictions)))
# Should be < 1e-6
```

A difference larger than `1e-4` indicates a coefficient or transformation
mismatch. Check variable names and ordering carefully.

---

## Step 4: Upload to GitHub

Upload the generated file to your private repository at the path:

```
<model_id>/<model_id>_specification.json
```

For example, for `model_id = "dvt_v1"`:

```
my-clinical-models/
└── dvt_v1/
    └── dvt_v1_specification.json
```

You can do this via the GitHub web interface (drag and drop into the correct
folder) or using the command line:

```bash
git add dvt_v1/dvt_v1_specification.json
git commit -m "Add DVT v1 model specification"
git push
```

---

## Step 5: Issue access tokens to evaluators

For each evaluator (or group of evaluators), create a fine-grained personal
access token:

1. Go to GitHub → Settings → Developer settings → Personal access tokens →
   Fine-grained tokens
2. Click **Generate new token**
3. Under **Repository access**, select **Only select repositories** and
   choose your model repository
4. Under **Permissions**, set **Contents** to **Read-only**
   (no other permissions are needed)
5. Set an appropriate **Expiration** — we recommend a short-lived token
6. Generate and copy the token

Send the token to the evaluator along with your GitHub username, repository
name, and model ID.

**Issue a unique token per evaluator.** This lets you revoke access for a
specific evaluator without affecting others and creates an audit trail of who
has accessed the model specification.

---

## Repository structure summary

```
my-clinical-models/                          ← private GitHub repository
├── dvt_v1/
│   └── dvt_v1_specification.json            ← model_id = "dvt_v1"
└── bc_cox_v1/
    └── bc_cox_v1_specification.json          ← model_id = "bc_cox_v1"
```

The evaluator calls:

```{r}
secure_model_validation(
  repo_owner      = "your-github-username",
  repo_name       = "my-clinical-models",
  model_id        = "dvt_v1",
  github_token    = "token_you_issued",
  validation_data = their_data,
  outcome         = "dvt"
)
```

---

## Checklist before sharing with evaluators

- [ ] `register_model()` ran without errors and printed a confirmation
      message
- [ ] Manual verification confirms predictions match those from the original
      model object (difference < 1e-4)
- [ ] Specification file is uploaded to the correct path:
      `<model_id>/<model_id>_specification.json`
- [ ] Repository visibility is set to **Private** if necessary
- [ ] Access token has read-only Contents scope and appropriate expiration
- [ ] Evaluator has been told:
  - Your GitHub username (`repo_owner`)
  - The repository name (`repo_name`)
  - The model ID (`model_id`)
  - Which variables their dataset must contain (`variables` field in the
    model description)
  - Any transformations they need to apply (if not embedded in preprocessing)
