---
title: "Guide for Evaluators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guide for Evaluators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

The **evaluatr** package allows you to independently evaluate a clinical
prediction model in your own dataset, without the model developer ever seeing
your patient data. This guide walks you through the complete workflow from installation to performance
reporting.

The system works as follows. The model developer stores their model
specification in a private GitHub repository. You use evaluatr to retrieve those specifications and compute predictions entirely within your own R session. Your patient data never leaves your computer.

## Installation

```{r}
devtools::install_github("JoieEnsor/evaluatr")
library(evaluatr)
```

## Finding available models

Use `list_registered_models()` to browse publicly registered models and find
the developer contact for requesting access:

```{r}
models <- list_registered_models()
print(models[, c("model_id", "model_name", "developer_email")])
```

## What the developer will send you

Once a developer has agreed to grant you access, they will send you four
pieces of information:

1. Their **GitHub username** — identifies the account hosting the model
2. The **repository name** — the name of the private repository
3. The **model ID** — a short identifier matching a folder in the repository
   (e.g. `"dvt_model_v1"`)
4. A **time-limited access token** — a fine-grained GitHub personal access
   token giving read-only access to their repository

Pass the token directly into the function call as shown in the examples below.

---

## Example 1: Basic whole-dataset evaluation

This example uses the DVT diagnosis dataset included with the package,
validated against a logistic regression model hosted on the public evaluatr
demo repository. It illustrates the complete evaluatr workflow for a binary
outcome model, and you can run it yourself exactly as written — no developer
contact required.

```{r}
library(evaluatr)

# Load the DVT example dataset
dvt_path <- system.file("extdata", "dvt_example.csv", package = "evaluatr")
dvt_data <- read.csv(dvt_path)

# Use studies 6 and 7 as the external validation set
# (studies 1-5 were used for model development)
dvt_val <- dvt_data[dvt_data$study %in% c(6, 7), ]

# Run secure validation against the public evaluatr demo model.
# The token below is a shared public demo credential (not a real GitHub
# token) for demonstration purposes only. In real use, the model developer
# provides you with a private-repo GitHub token.
result <- secure_model_validation(
  repo_owner      = "JoieEnsor",
  repo_name       = "evaluatr-demo-models",
  model_id        = "dvt_logistic_v1",
  github_token    = "evaluatr-demo",
  validation_data = dvt_val,
  outcome         = "dvt"
)
```

The `result` object contains a shuffled prediction-outcome matrix. The
shuffling means that predictions cannot be matched back to individual patient
rows, which ensures the package
cannot be used to generate predictions for individual patients in clinical
practice.

```{r}
# Compute performance metrics
performance <- eval_performance(
  validation_result    = result,
  generate_plots       = TRUE,
  confidence_intervals = TRUE,
  n_boot               = 50,      # low here for illustration purposes
  decision_threshold   = 0.15     # set to the clinically relevant threshold
)

# View the metrics table
performance$metrics

# Access individual diagnostic plots
performance$plots$calibration      # calibration plot
performance$plots$decision_curve   # decision curve analysis
performance$plots$distribution     # predicted probability distributions
```

The metrics table includes the following:

| Domain | Metrics |
|--------|---------|
| Discrimination | AUROC / c-statistic |
| Calibration | Intercept, slope, ICI, ECI, ECE, O:E ratio |
| Overall performance | Brier score, Nagelkerke R², Cox-Snell R², McFadden R² |
| Clinical utility | Net benefit, standardised net benefit |

---

## Example 2: Subgroup evaluation

Many validation studies assess model performance separately within clinically
relevant subgroups, such as sex, age group, or recruitment site. evaluatr
supports this through the `by` argument.

The workflow has two stages. First, pass `by` to `secure_model_validation()`
so that the subgroup label is retained alongside each shuffled prediction.
Second, pass `result$shuffled_by` to `eval_performance()` to obtain separate
metrics per subgroup.

```{r}
# Stage 1 — run validation retaining study labels
result_by_study <- secure_model_validation(
  repo_owner      = "developer-username",
  repo_name       = "my-models",
  model_id        = "dvt_v1",
  github_token    = "token_provided_by_developer",
  validation_data = dvt_val,
  outcome         = "dvt",
  by              = "study"     # column name in your dataset
)

# Stage 2 — compute metrics separately per study
performance_by_study <- eval_performance(
  validation_result    = result_by_study,
  generate_plots       = TRUE,
  confidence_intervals = TRUE,
  n_boot               = 50,      # low here for illustration purposes
  decision_threshold   = 0.15,
  by                   = result_by_study$shuffled_by
)

# The metrics table now has a Subgroup column
performance_by_study$metrics

# Plots are organised by subgroup level
performance_by_study$plots[["6"]]$calibration   # study 6
performance_by_study$plots[["7"]]$calibration   # study 7
```

Each subgroup must contain at least 50 observations. If any subgroup is
smaller than this, `secure_model_validation()` will return an informative
error before any predictions are computed.

---

## Example 3: Models requiring pre-processed variables

Some models are developed using transformed or derived variables — for
example, an age-squared term or a log-transformed biomarker. How you handle
this depends on whether the developer has embedded the transformation in their
model specification.

### When you need to pre-process (developer specifies transformed variable names)

If the model requires a variable named `log_pr` (log-progesterone), you will
need to create that column in your dataset before calling
`secure_model_validation()`. The developer should document which
transformations are required in their publication or access token
communication.

```{r}
bc_path <- system.file("extdata", "bc_survival_example.csv",
                       package = "evaluatr")
bc_val  <- read.csv(bc_path)
bc_val  <- bc_val[bc_val$val == 1, ]

# Create the required derived variable
bc_val$log_pr <- log(bc_val$pr + 1)

# Now run validation
result_bc <- secure_model_validation(
  repo_owner      = "developer-username",
  repo_name       = "my-models",
  model_id        = "bc_cox_v1",
  github_token    = "token_provided_by_developer",
  validation_data = bc_val,
  outcome         = "rfi"
)
```

### When the developer handles pre-processing automatically

Developers can embed transformation code directly in their model
specification. In this case, you provide the raw data and evaluatr handles
the rest — no additional steps are required on your side.

```{r}
# No pre-processing needed — the developer's specification handles it
result_autoprepro <- secure_model_validation(
  repo_owner      = "developer-username",
  repo_name       = "my-models",
  model_id        = "bc_cox_auto_v1",
  github_token    = "token_provided_by_developer",
  validation_data = bc_val,    # raw pr provided as-is
  outcome         = "rfi"
)
```

The two approaches produce identical predictions; the difference is purely
in who applies the transformation.

---

## More examples: other outcome types

The examples above focus on binary outcomes and general workflow concepts.
The public evaluatr demo repository also hosts models covering the other
outcome types evaluatr supports. Runnable scripts for each are included with
the package under `system.file("extdata", package = "evaluatr")`:

| Script | Model type | Outcome |
|--------|-----------|---------|
| `example_01_dvt_logistic.R` | Logistic regression | Binary (DVT diagnosis) |
| `example_02_bc_cox.R` | Cox proportional hazards | Survival (breast cancer relapse) |
| `example_03_bc_weibull.R` | Weibull AFT | Survival (breast cancer relapse) |
| `example_04_adnex_multinomial.R` | Multinomial logistic regression | 5-category (ADNEX ovarian mass classification) |

Each script uses the same public demo repository and read-only token as
Example 1 above, so they can be run as-is. The survival and multinomial
scripts also include independent manual verification of the predictions
against a locally-fit reference model.

---

## Minimum dataset requirements

| Requirement | Value |
|-------------|-------|
| Minimum total observations | 50 |
| Minimum per subgroup (when using `by`) | 50 |
| Outcome variable type for `eval_performance()` | Binary (0/1) in the current release. Survival and multinomial predictions can be computed with evaluatr, but performance metrics for these outcome types must currently be calculated independently by the evaluator; full support in `eval_performance()` is planned for a future release. |

These thresholds exist to prevent the package from being used to obtain
predictions for individual patients.

---

## Security and data privacy

- **Your patient data never leaves your R session.** No data is transmitted
  to GitHub or to the model developer.
- **Predictions are shuffled** before being returned, so they cannot be
  matched back to individual patient rows.
- **Every validation run is logged** by the evaluatr registry, creating an
  audit trail. Access tokens can be revoked by the developer at any time.
- The package is designed exclusively for **retrospective evaluation
  studies**. It must not be used to generate predictions that inform care
  for individual patients.

---

## Getting help

If you encounter problems, check the following:

- The token provided by the developer has not expired.
- Your dataset contains all the variable names listed in the developer's
  documentation.
- If using `eval_performance()`, check that the outcome variable is coded
  as 0/1 for binary outcomes.
- The `model_id`, `repo_owner`, and `repo_name` all match exactly what the
  developer provided (these are case-sensitive).

See `?secure_model_validation` and `?eval_performance` for full argument
documentation.
