---
title: "Getting Started with evaluatr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with evaluatr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

The **evaluatr** package enables independent external validation of clinical
prediction models while preserving model confidentiality and patient data
privacy. It implements a two-component system:

- **Developer side:** model coefficients and prediction code are stored in a
  private GitHub repository.
- **Evaluator side:** this package retrieves and uses those specifications
  locally, without exposing coefficients or transmitting patient data.

## Installation

```{r}
# Install from GitHub (requires devtools)
devtools::install_github("JoieEnsor/evaluatr")
```

## Step 1 — Obtain access from the model developer

The model developer will provide you with:

1. Their **GitHub username** (e.g. `"developer-name"`)
2. The **repository name** (e.g. `"my-models"`)
3. The **model ID** (e.g. `"aki_model_v1"`) — this identifies the folder
   within the repository
4. A **time-limited fine-grained GitHub PAT** with read-only access to that
   repository

Store the token securely (e.g. with `usethis::edit_r_environ()`):

```{r}
# ~/.Renviron
EVALUATR_TOKEN=github_pat_...
```

## Step 2 — Run secure validation

```{r}
library(evaluatr)

# Load your validation dataset
my_data <- read.csv("path/to/my_validation_data.csv")

# Compute predictions
result <- secure_model_validation(
  repo_owner      = "developer-name",
  repo_name       = "my-models",
  model_id        = "aki_model_v1",
  github_token    = Sys.getenv("EVALUATR_TOKEN"),
  validation_data = my_data,
  outcome         = "AKI"   # name of outcome column in my_data
)
```

The returned `result` object contains a **shuffled** prediction–outcome
matrix. Individual predictions cannot be traced back to patient rows.

### Subgroup analysis

```{r}
result_by_sex <- secure_model_validation(
  repo_owner      = "developer-name",
  repo_name       = "my-models",
  model_id        = "aki_model_v1",
  github_token    = Sys.getenv("EVALUATR_TOKEN"),
  validation_data = my_data,
  outcome         = "AKI",
  by              = "sex"   # subgroup variable name
)
```

## Step 3 — Compute performance metrics

```{r}
perf <- calculate_pmextval_metrics(
  validation_result  = result,
  generate_plots     = TRUE,
  confidence_intervals = TRUE,
  n_boot             = 500,
  decision_threshold = 0.25
)

# View the metrics table
perf$metrics

# Access individual plots
perf$plots$calibration
perf$plots$decision_curve
perf$plots$distribution
```

### Subgroup metrics

```{r}
perf_by_sex <- calculate_pmextval_metrics(
  validation_result  = result_by_sex,
  generate_plots     = TRUE,
  confidence_intervals = TRUE,
  n_boot             = 500,
  decision_threshold = 0.25,
  by                 = result_by_sex$shuffled_by
)

perf_by_sex$metrics   # includes a 'Subgroup' column
```

## Metrics computed

| Domain | Metric |
|--------|--------|
| Discrimination | AUROC / c-statistic |
| Calibration | Intercept, Slope, ICI, ECI, ECE, O:E ratio |
| Overall | Brier score, Nagelkerke R², Cox–Snell R², McFadden R² |
| Utility | Net benefit, Standardised net benefit |

## Security notes

- Model coefficients are used in-memory only and immediately cleared after
  predictions are computed.
- Patient data never leaves your local environment.
- The shuffled output matrix prevents coefficient reverse-engineering.
- A minimum of 20 observations per (sub)group is enforced.
- The package is designed for **evaluation purposes only**, not for making
  predictions to guide individual patient care.
