---
title: "Guide for Model Developers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guide for Model Developers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

This guide explains how to set up your model so that independent evaluators can use the evaluatr package to validate it in their own datasets, without you ever accessing their patient data and without them ever seeing your model coefficients.

The system requires two things from you as a developer:

1. A **private GitHub repository** containing a JSON file with your model specification
2. A **fine-grained access token** that you issue to each evaluator, giving them read-only access to your repository

This takes approximately 30 minutes to set up for the first model, and much less for subsequent ones.

---

## Step 1: Create a private GitHub repository

If you do not already have a GitHub account, create one at [github.com](https://github.com). Then create a new repository:

- Set the visibility to **Private**
- Choose a clear, memorable name (e.g. `my-clinical-models`)
- Do not initialise with a README

Your repository will hold one or more model specifications. Each model lives in its own subfolder following this structure:

```
my-clinical-models/
└── models/
    ├── model_v1/
    │   └── coefficients.json
    └── model_v2/
        └── coefficients.json
```

The folder name (e.g. `model_v1`) becomes the `model_id` that evaluators will use. The file must be named `coefficients.json`.

---

## Step 2: Write your model specification JSON

The JSON file contains four fields: `model_type`, `coefficients`, `prediction_function`, and `metadata`. The examples below cover the most common model types.

### Example A: Standard logistic regression

This is the simplest case. The coefficients are named to match the variable names in the development dataset, and the prediction function converts the linear predictor (LP) to a probability.

The LP is computed automatically by evaluatr using the named coefficients and the evaluator's data — you do not need to compute it yourself in the prediction function for standard models.

```json
{
  "model_type": "logistic_regression",
  "coefficients": {
    "(Intercept)": -1.25,
    "age": 0.02,
    "biomarker_score": 0.80,
    "treatment_group": -0.60
  },
  "prediction_function": "probabilities <- 1 / (1 + exp(-LP)) \n result_matrix <- matrix(probabilities, ncol = 1) \n colnames(result_matrix) <- 'probability' \n result_matrix",
  "required_packages": [],
  "metadata": {
    "model_name": "My Clinical Risk Score",
    "version": "1.0",
    "created_date": "2025-01-01",
    "outcome_type": "binary",
    "variables": ["age", "biomarker_score", "treatment_group"],
    "description": "Logistic regression predicting 30-day readmission."
  }
}
```

The `prediction_function` field contains R code as a single string. Each new line in your R code is represented by `\n`. The code has access to `LP` (the pre-computed linear predictor) and must return a matrix. See the formatting section below for how to write this comfortably.

### Example B: Model using transformed variables — evaluator pre-processes

If your model uses a transformed predictor (e.g. log-creatinine), you have two options. In this example, the evaluator is expected to create the transformed column themselves before calling `secure_model_validation()`. The `variables` field lists the transformed name, signalling to the evaluator what column must exist.

```json
{
  "model_type": "logistic_regression",
  "coefficients": {
    "(Intercept)": 2.10,
    "age": 0.018,
    "log_creatinine": 1.89
  },
  "prediction_function": "probabilities <- 1 / (1 + exp(-LP)) \n result_matrix <- matrix(probabilities, ncol = 1) \n colnames(result_matrix) <- 'probability' \n result_matrix",
  "required_packages": [],
  "metadata": {
    "model_name": "AKI Risk Score (log creatinine)",
    "version": "1.0",
    "created_date": "2025-01-01",
    "outcome_type": "binary",
    "variables": ["age", "log_creatinine"],
    "description": "Requires log_creatinine = log(creatinine + 0.001) to be created by the evaluator before validation."
  }
}
```

Always document the exact transformation required in your model description and in your published paper. Evaluators will create the column and pass the dataset with the transformed variable included.

### Example C: Model using transformed variables — developer embeds pre-processing

Alternatively, you can embed the transformation directly in the `prediction_function`. This is the recommended approach when possible: the evaluator provides raw data and your code handles everything. This eliminates the risk of the evaluator applying a slightly different transformation.

When you embed pre-processing, you must also recompute the LP yourself inside the prediction function (rather than relying on the automatically computed LP), because the transformation creates the new variable after the automatic LP computation has already run.

```json
{
  "model_type": "logistic_regression",
  "coefficients": {
    "(Intercept)": 2.10,
    "age": 0.018,
    "log_creatinine": 1.89
  },
  "prediction_function": "validation_data$log_creatinine <- log(validation_data$creatinine + 0.001) \n LP <- `(Intercept)` + age * validation_data$age + log_creatinine * validation_data$log_creatinine \n probabilities <- 1 / (1 + exp(-LP)) \n result_matrix <- matrix(probabilities, ncol = 1) \n colnames(result_matrix) <- 'probability' \n result_matrix",
  "required_packages": [],
  "metadata": {
    "model_name": "AKI Risk Score (log creatinine, auto pre-processing)",
    "version": "1.0",
    "created_date": "2025-01-01",
    "outcome_type": "binary",
    "variables": ["age", "creatinine"],
    "description": "Evaluator provides raw creatinine. Log transformation is applied automatically."
  }
}
```

Note that the `variables` field now lists `creatinine` (the raw column the evaluator provides), not `log_creatinine`. This is what evaluatr checks for when verifying the evaluator's dataset contains the required columns.

In the prediction function, the coefficient names (e.g. `log_creatinine`, `age`) are available as named objects in the execution environment alongside `validation_data`. This is why `log_creatinine * validation_data$log_creatinine` correctly multiplies the coefficient value by the data column.

### Example D: Multinomial logistic regression

For multinomial models, the prediction function must return a matrix with one column per outcome category. LP is not pre-computed by evaluatr for multinomial models — your prediction function handles everything.

```json
{
  "model_type": "multinomial_logistic",
  "coefficients": {
    "z1_intercept": -3.46,
    "z1_age": 0.012,
    "z2_intercept": -3.36,
    "z2_age": 0.011
  },
  "prediction_function": "z1 <- z1_intercept + z1_age * validation_data$age \n z2 <- z2_intercept + z2_age * validation_data$age \n denom <- 1 + exp(z1) + exp(z2) \n result_matrix <- cbind( \n  category_A = 1 / denom, \n  category_B = exp(z1) / denom, \n  category_C = exp(z2) / denom \n ) \n result_matrix",
  "required_packages": [],
  "metadata": {
    "model_name": "Three-Category Outcome Model",
    "version": "1.0",
    "created_date": "2025-01-01",
    "outcome_type": "multinomial",
    "outcome_categories": ["category_A", "category_B", "category_C"],
    "reference_category": "category_A",
    "variables": ["age"],
    "description": "Multinomial logistic regression with three outcome categories."
  }
}
```

---

## Writing the prediction_function field comfortably

The JSON format requires the prediction function to be a single string with `\n` representing line breaks. Writing this directly in a JSON file is awkward. The recommended workflow is:

1. Write and test your prediction function as a normal R script
2. When satisfied, convert it for JSON by replacing each newline with `\n`

You can do this conversion in R:

```{r}
# Write your function as normal R code
my_function <- '
probabilities <- 1 / (1 + exp(-LP))
result_matrix <- matrix(probabilities, ncol = 1)
colnames(result_matrix) <- "probability"
result_matrix
'

# Convert to JSON-safe single string
json_safe <- gsub("\n", "\\n", trimws(my_function))
cat(json_safe)
# Copy the output and paste into your JSON file
```

A developer utility function to generate complete JSON files from R model objects is planned for a future version of evaluatr.

---

## Step 3: Verify your specification

Before sharing with any evaluator, test that your JSON produces correct predictions. The cleanest approach is to apply your coefficients manually in R and compare to what evaluatr returns.

```{r}
library(evaluatr)

# Use your own development dataset for this check
dev_data <- read.csv("path/to/your/development_data.csv")

# Compute predictions manually (exactly as you would from the model object)
manual_lp    <- -1.25 + 0.02 * dev_data$age + 0.80 * dev_data$biomarker_score
manual_probs <- 1 / (1 + exp(-manual_lp))

# Compute predictions via evaluatr using your GitHub repo
result <- secure_model_validation(
  repo_owner      = "your-github-username",
  repo_name       = "my-clinical-models",
  model_id        = "model_v1",
  github_token    = Sys.getenv("GITHUB_PAT"),   # your own PAT, full access
  validation_data = dev_data,
  outcome         = "outcome_column"
)

# Sort both sets of predictions and compare
# (result is shuffled, so sort before comparing)
max(abs(sort(manual_probs) - sort(result$shuffled_predictions)))
# Should be < 1e-10 if the JSON is correct
```

If the maximum difference exceeds `1e-10`, there is a precision or formula error in your JSON. Check coefficient values and the prediction function code carefully.

---

## Step 4: Issue access tokens to evaluators

For each evaluator (or group of evaluators), create a fine-grained personal access token:

1. Go to GitHub → Settings → Developer settings → Personal access tokens → Fine-grained tokens
2. Click **Generate new token**
3. Under **Repository access**, select **Only select repositories** and choose your model repository
4. Under **Permissions**, set **Contents** to **Read-only** (no other permissions needed)
5. Set an appropriate **Expiration** — we recommend a short-lived token
6. Generate and copy the token

Send the token to the evaluator along with your GitHub username, repository name, and model ID. The token gives read-only access to the repository contents only — it cannot be used to modify your repository or access any other part of your GitHub account.

**Each evaluator should receive a unique token.** This allows you to revoke access for a specific evaluator without affecting others, and gives you an audit trail of who has accessed the model specification.

---

## Repository structure summary

```
my-clinical-models/               ← private GitHub repository
└── models/
    ├── model_v1/
    │   └── coefficients.json     ← model_id = "model_v1"
    └── model_v2/
        └── coefficients.json     ← model_id = "model_v2"
```

The evaluator calls:

```{r}
secure_model_validation(
  repo_owner   = "your-github-username",
  repo_name    = "my-clinical-models",
  model_id     = "model_v1",
  github_token = "token_you_issued",
  ...
)
```

---

## Checklist before sharing with evaluators

- [ ] JSON file is valid (can paste into [jsonlint.com](https://jsonlint.com) to check)
- [ ] Coefficient names in JSON match variable names in the development model
- [ ] `variables` field in metadata lists all required predictor column names
- [ ] Prediction function returns a matrix with named columns
- [ ] Manual verification confirms predictions match those from the original model object
- [ ] Access token issued with read-only scope and appropriate expiration
- [ ] Evaluator has been told which variables to prepare (including any transformations if using Example B approach)
